Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning

### 2022-08-08 manuscript/manuscript.Rmd
Little edit on general ml paragraph
Some more touch ups

Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis_statistical_2018].
Deep learning models are often used to make point forecasts, but for their application to ecological time series, it will often be necessary to use multi-step, probabilistic forecasts. 
For all the deep learning models in this study, we use the same general process.
Each machine learning model is trained on one time series.
For the Hopf and saddle node cases, these time series consist of the period leading up to the bifurcation.
A critical transition is, however, included in the training set for the stochastic transition case.
Each model is trained to learn the parameters of a Laplace distribution for every time step in the forecast horizon.
To produce a forecast, we input a time series into a model, then we draw samples from the distributions that were learned during training.

@@

The MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
This is a much stronger assumption than that made by the ML models, though it can potentially be justified on the basis of a mechanistic understanding of the processes involved.
It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994].
Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
Because the available data come from only a small region of the dynamical state space, they are consistent with many possible parameterizations of the same model structure -- which creates likelihood ridges and non-identifiability of specific parameter values.
Using more simplified versions of the dynamical processes in question, such as the canonical form of a bifurcation, can mitigate this issue in some cases.
Even when such non-identifiability issues cannot be avoided entirely, they can usually be diagnosed by examining the degree of mixing in MCMC sampling and comparing posterior to prior distributions.
<!-- While similar convergence issues may arise in the estimation of ML models, they may be harder to diagnose. -->

@@

There are a number of questions that we have left unanswered that we hope will be addressed in future work.
In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties. <!-- MFL: I think we could cut this paragraph out completely now -->

>>>>

Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis_statistical_2018].
Deep learning models are often used to make point forecasts, but for their application to ecological time series, it will often be necessary to use multi-step, probabilistic forecasts. 
For all the deep learning models in this study, we use the same general process.
Each machine learning model is trained on one time series drawn from the three scenarios described previously.
For the Hopf and saddle node cases, these time series consist of the period leading up to the bifurcation.
A critical transition is, however, included in the training set for the stochastic transition case.
Each model is trained to learn the parameters of a Laplace distribution for every time step in the forecast horizon.
To produce a forecast, we input a time series into a model, then we draw samples from the distributions that were learned during training.

@@

The MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
This is a much stronger assumption than that made by the ML models, though this assumption can potentially be justified on the basis of a mechanistic understanding of the processes involved.
It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994]. <!-- ML: maybe "In the real world, having access to the _true_ model is never the case:"-->
Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
Because the available data come from only a small region of the dynamical state space, they are consistent with many possible parameterizations of the same model structure -- which creates likelihood ridges and non-identifiability of specific parameter values.
Using more simplified versions of the dynamical processes in question, such as the canonical form of a bifurcation, can mitigate this issue in some cases.
Even when such non-identifiability issues cannot be avoided entirely, they can usually be diagnosed by examining the degree of mixing in MCMC sampling and comparing posterior to prior distributions.
<!-- While similar convergence issues may arise in the estimation of ML models, they may be harder to diagnose. -->

@@

There are a number of questions that we have left unanswered that we hope will be addressed in future work.
In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data or observation error.
